
---
title: "Basic of the R lenguage"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 12, fig.height = 4)
```



## Main types of objects

Vectors are most common and basic data structure in R and is pretty much the workhorse of R. A vector is a collection of elements that are most commonly of mode `character`, `logical`, `integer` or `numeric`. 

Here are some vectors (length one) of different classes: 

```{r}

# Character
a_caracther <- "Hello"
a_caracther
class(a_caracther)

# Numeric
a_number   <- 3; a_number
4 -> another_number; another_number

# Also numeric but infinite 
neg_inf <- log(0); neg_inf
pos_inf <- 1/0; pos_inf


# Logical vectors

a_truth <- TRUE; a_truth

```
<br>


### Vectors

Usually vectors (of any type) have a length greater than 1. Importantly, in R the elements of a vectors must be of the same data type. Otherwise all elements will be coerced to `character`.

To input these vectors manually you will need to use the operator `c()`, which stands for "concatenate". For example: 


```{r}

# numeric
u <- c(1,2,3,4,5,6); u
v <- c(7,8,9,10,11); v
uv <- c(u,v); uv

# characters

some_letters  <- c("a","b")
more_letters  <- c("c","d") 
abcd  <- c(some_letters,more_letters); abcd

```
<br>

There are many other more convenient ways to create vectors. For example:

```{r}

# Repeat value 1, six times
w <- rep(1,6); w

# Sequence starting in 1, ending in 2, spaced by 0.5
x <- seq(from=1,to=3.5,by=0.5); x

# Sequence of length 6, starting in 10 and ending in 10, spacing authomatically determined.
y <- seq(length=6, from=10, 100); y

# Shortcut for seq(6,11,1)
z <- 6:11; z

# You can combine these functions at will. Example: create a vector that repeats x twice

xx <- rep(x,2); xx

```
<br>


### Matrices

In R matrices are an extension of vectors. They are not a separate type of object but vector with dimensions: the number of rows and columns. Again, matrices can be created in many ways.

```{r}

# Input manually 

some_numbers <- seq(1:100)

a_matrix <- matrix(some_numbers,  nrow = 2, ncol = 2); a_matrix
another_matrix <- matrix(some_numbers,  nrow = 5); another_matrix

# Binding vectors, eithers by rows or by columns

matrix_1 <- cbind(y,x); matrix_1
matrix_2 <- rbind(y,x); matrix_2

# Other special cases: a 10 by 10 identity matrix 

identity <- diag(10); identity

```
<br>


### Data Frames

A data frame is a very important data type in R. It’ the structure for most data we use for statistics and the default object in software such as `Stata`, `SAS`, `SPSS`, etc. Unlike vectors and matrices, the content of a data frame is not restricted to a single type and can encompass any mixture of data types. As in matrices, however, in a data frame every element (or "variable) has to have same length (i.e. same number of "observations").



```{r}

# create a data frame
mydata <- data.frame(id = letters[1:10], x = 1:10, y = 11:20); mydata

# access variables in a data frame
mydata$x

# assign a new element to a data frame

mydata$z <- mydata$x + mydata$y; mydata

```
<br>


### Lists

R lists work as containers. They can contain any type of data, of any lengh. A list can even contain other lists.

```{r}

my_list <- list(myvector=letters[1:10], mymatrix=matrix(1:100,nrow=5), mydf=mydata); my_list

my_list$mymatrix 
```
<br>


### Converting objects

Object of one type can be converted into objects of a different type using the functions of the type `as.___()`. Example:

```{r,error=TRUE}

# matrix to data.frame
df_2 <- as.data.frame(matrix_2)
names(df_2) <- c("var1","var2")

# character to numeric
as.numeric("3") + 1

# numeric to character
as.character(3)
```
<br>

## Operations 


### Basic operations with text

```{r}

greetings <- paste("Hello","Word",sep=" "); greetings

five_letters <- paste(c("a","b","c","d"), collapse="+"); five_letters

```
<br>


### Arithmetic 

```{r}

addition    <- 1 + 2
substaction <- 1 - 2 
product     <- 1*2
ratio       <- 1/2
power       <- 1^3
root        <- 1^(1/3)
natlog      <- log(1)
rounded     <- round(root,1)
```
<br>

### Computing quantities of interest from vectors. 

For this example we will generate two vectors using a random number generator. In each case we are going to draw a sample of 20 values from a [uniforme distribution](https://en.wikipedia.org/wiki/Uniform_distribution_(continuous)) between 0 and 100. 

```{r}

x <- round(runif(20,0,100),0)
y <- round(runif(20,0,100),0); y

# We can compute many quantities of interest. Here some examples

# number of rows
n <- length(y); n

# sum of entries
sum_y <- sum(y); sum_y

# value of smallest entry
min_y <- min(y); min_y

# position of smallest entry
index_min_y <- which.min(y); index_min_y

# value of largest entry
max_y <- max(y)

# position of largest entry
index_max_y <- which.max(y)

# Returns the a ranking of values
rank_y <- rank(y); rank_y

# Returns a vector of sorted values
sort_y <- sort(y); sort_y

```
<br>


### Vector operations

You can do mathematical manipular of vectors and matrices as long as the dimensions are right.
All arithmetic operations on vectors and matrices are interpreted as "element wise" operations. For vector/matrix multiplication (i.e. [dot products](https://en.wikipedia.org/wiki/Dot_product), [outer products](https://en.wikipedia.org/wiki/Outer_product), [matrix multiplication](https://en.wikipedia.org/wiki/Matrix_multiplication)) you need to use `%*%` instead of `*`. Example:

```{r}


# 2 vectors of same dimensions
x <- round(runif(20,0,100),0)
y <- round(runif(20,0,100),0)

# 2 matrices of same dimensions

A <- matrix(round(runif(100,0,100),0), nrow=10)
B <- matrix(round(runif(100,0,100),0), nrow=10); B


# Linear combination of vectors

lincom_vectors <- 2*x + y; lincom_vectors 


# Element wise operations

x_times_y <- x*y; x_times_y

x_divided_y <- x/y; x_divided_y

A_times_B <- A*B; A_times_B

A_divided_B <- A/B; A_divided_B


#  vector multiplication

dot_product <- t(x)%*%y; dot_product 

outer_product <- x%*%t(y); outer_product

# Matrix multiplication 

AB <- A%*%B; AB
BA <- B%*%A; BA

```
<br>

## Logical operations

Logical operators evaluate of truth value of an statement. They return a decision in terms of TRUE or FALSE.


```{r}

# The following are the main logical operators:# <	less than
# <=	less than or equal to
# >	greater than
# >=	greater than or equal to
# ==	exactly equal to
# !=	not equal to
# !x	Not x
# x | y	x OR y
# x & y	x AND y
# isTRUE(x)	test if X is TRUE

#These operators act on raw, logical and number-like vectors.

y; y>20

(y>20)*1

round(y,0)==round(x,0)

y!=x

```
<br>

### Subseting

There are many methods to access a subset of data, be it a vector, matrix, data frame or list.

```{r,error=TRUE}

# Indexing a vector

y[1]
y(1)

y[c(1,3)]
y[3:length(y)]

# Negative indexing in vectos

y[-c(1,3)]

# Indexing by names

names(y) <- paste("Name", 1:length(y), sep = "")
y
y[c("Name1","Name5","Name12")]

# Indexing in matrices

M_2[2,5]
M_2[,5]
M_2[1,]


# Subseting using logical conditions

y[which(y<2 | y>=5)]

y[y<2 | y>=5]

# Assign and change values  using logic operators

y[y<2 | y>=5] <- 0

```
<br>


## Missing values 

In R, missing values are represented by the symbol NA (not available). Impossible values (e.g., dividing by zero) are represented by the symbol NaN (not a number). 

```{r}

# Data with missing values
y <- c(3,1,NA,4,6,NA)

# Return logical vectors
is.na(y)

# Replacing a missing value
y[is.na(y)==TRUE] <- 1

```
<br>

## Functions 

One of the great strengths of R is the user's ability to add functions. In fact, many of the functions in R are actually functions of functions. An example of a function is given below.

```{r}

my_function <- function(x) {
		fx <- log(x + (x^2)/2 + (x^3)/3 + (x^4)/4 + (x^5)/5)
		return(fx)
	}

my_function(1)
my_function(2)
```
<br>


## Loops

Loops are used in programming to repeat a specific block of code. 

### For loops

A `for loop` is used to iterate over a list of values (a vectors). Example

```{r}

# For loop on list of characters
list_words <- c("Welcome", "to", "the", "Introduction","to","R")

for (i in list_words) {
	print(i)
}
```
<br>


```{r}

# For loop on list of numbers
for (i in 1:50) {
	r <- my_function(i)
	print(r) 
}
```
<br>


```{r}

my_function(1:50)

```
<br>

####  While loops
`While loops` are used to loop until a specific condition is met. Example

```{r}

# start on value i=1
i=1
trace <- NULL

# redefine i such i equals i + function(i). Stop when i take on value 100 or more.

while (i < 100) {
  trace <- rbind(trace,c(i,i + my_function(i)))
	i = i + my_function(i)
}

trace
```
<br>


## Plot 

```{r}

x <- 1:50
y <- my_function(1:50)


plot(x,y, col="red", ylab="log(x + x^2/2 + ... + x^5/5", main="My first function!")
```
<br>